% !Mode:: "TeX:UTF-8"


\begin{chineseabstract}

近年来，人口老龄化问题不断加深，慢性病人群规模持续扩大，关于慢性病的相关健康咨询需求不断增加；然而，传统线下问诊在便利性和及时性方面存在一定限制，难以满足用户日常持续性的健康信息获取需求。与此同时，大语言模型在自然语言理解和生成任务中展现出了较强的能力，为智能问答系统在医疗健康领域的应用提供了新的技术路径。

但是，在医学等高风险应用场景中，大语言模型仍存在一定局限，例如会产生事实性错误（幻觉）、安全性约束不足以及专业知识的理解不足。因此，有必要结合外部知识与模型优化方法，对模型进行优化，以提升其在医学场景中的安全性与可靠性。

针对上述问题，本文围绕“三高”慢性病的咨询场景，设计并实现基于检索增强生成的医学问答系统。本文通过国家政策、学术期刊和医学指南等可靠来源搜集相关资料，构建医学知识库，并结合语义检索和关键词检索精准匹配医学文档，从而为大语言模型提供外部知识支撑。针对模型安全性约束不足等问题，本文引入人类偏好优化方法，通过偏好数据对模型进行微调，使其生成内容更符合医学场景下的安全性和规范性要求。在此基础上，本文完成了医学问答系统的整体设计与实现，实现了从知识构建、检索增强到模型生成的完整问答流程。

实验结果表明，混合检索在召回覆盖率上优于单一路径检索，Recall@5召回率达到89\%；同时，基于DPO优化后的模型在盲评中相较于基础模型取得59.2\%的胜率，优化后的模型在事实一致性和完备性上均优于基础模型。

\chinesekeyword{检索增强生成；大语言模型；混合检索；直接偏好优化；医学问答系统}
\end{chineseabstract}

\begin{englishabstract}
In recent years, the problem of population aging has been deepening, the scale of people with chronic diseases has been expanding, and the demand for health consultation about chronic diseases has been increasing. However, the traditional offline consultation has some limitations in convenience and timeliness, and it is difficult to meet the daily and continuous health information acquisition needs of users. At the same time, the large language model shows strong ability in the task of natural language understanding and generation, which provides a new technical path for the application of intelligent question answering system in the medical and health field.

However, in high-risk application scenarios such as medicine, the big language model still has some limitations, such as factual errors (hallucinations), insufficient security constraints and insufficient understanding of professional knowledge. Therefore, it is necessary to combine external knowledge and model optimization methods to optimize the model to improve its safety and reliability in medical scenes.

In order to solve the above problems, this paper designs and implements a medical question answering system based on retrieval enhancement generation around the consultation scene of "three high" chronic diseases. The system collects relevant information from reliable sources such as national policies, academic journals and medical guides, constructs a medical knowledge base, and combines semantic retrieval and keyword retrieval to accurately match medical documents, thus providing external knowledge support for the large language model. In order to solve the problem of insufficient security constraints of the model, this paper introduces human preference optimization method, and fine-tunes the model through preference data to make its generated content more in line with the safety and normative requirements in medical scenes. On this basis, this paper completed the overall design and implementation of the medical question answering system, and realized the complete question answering process from knowledge construction, retrieval enhancement to model generation.

The experimental results show that hybrid retrieval is superior to single-path retrieval in recall coverage, and the recall rate of Recall@5 reaches 89\%. At the same time, compared with the basic model, the optimized model based on DPO achieved a winning rate of 59.2\% in blind evaluation, and the optimized model was superior to the basic model in fact consistency and completeness.

\englishkeyword{Retrieval-Augmented Generation; Large Language Model; Hybrid Retrieval; Direct Preference Optimization; Medical Question Answering System}
\end{englishabstract}
